---
title: "PS270 Final Project"
author: "Jared Haberman"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

I am interested in exploring data related to gerrymandering and voting rights. 

Does increasing the availability of ballot drop boxes increase voter participation in elections? If ballot drop boxes are more available for voters, then there will by higher voter turnout. The variable of interest is in state laws regarding ballot drop boxes. States can either require ballot drop boxes, allow them but not require them, have no law on drop boxes, restrict the use of ballot drop boxes, or ban their use altogether. I will assign each of these options a number 1-5, to measure the variable numerically. The outcome variable of interest is voter turnout. It is measured by the percentage of eligible voters in the state that voted in the 2024 election. If states that require ballot drop boxes have higher voter turnout, then my hypothesis will be supported. However, if states with more restrictive use on ballot drop boxes have higher voter turnout, or there is no correlation between ballot drop boxes and voter turnout, then my hypothesis will be disproved. 


```{r}
library(tidyverse)
```

```{r}
ballot <- read.csv("Final Project Data - Sheet1.csv")
glimpse(ballot)
```

```{r}
library(dplyr)

ballot <- ballot |>
  mutate(
    Voter_Participation_2024 = as.numeric(gsub("%", "", Voter_Participation_2024))
  )
ballot
```


```{r}
ballot_summary <- ballot |>
  group_by(Ballot_Box_Law) |>
  summarize(
    avg_turnout = mean(Voter_Participation_2024, na.rm = TRUE),
    .groups = "drop"
  )

ballot_summary

```

```{r}
Ballot_Plot <- ballot_summary |>
  ggplot(mapping = aes(x = fct_reorder(Ballot_Box_Law, avg_turnout),
                       y = avg_turnout)) +
  geom_bar(stat = "identity") +
  labs(x = "Ballot Box Law",
       y = "Avg Voter Turnout",
       title = "How Ballot Box Laws Effect Voter Turnout")
Ballot_Plot
```

```{r}
ballot_summary_table <- ballot_summary |>
  rename("Ballot Box Availability" = Ballot_Box_Law,
         "Average Voter Turnout" = avg_turnout)

knitr::kable(ballot_summary_table, digits = 2)
```

This data shows a somewhat small, but noticeable effect of ballot box availability on voter turnout. The more supportive a law is regarding ballot boxes, the higher voter turnout is. In states where ballot boxes are banned, voter turnout is about 61.3%, while in states where ballot boxes are required, voter turnout is at 69.7%. The rest of the availability options follow this trend as well, with restricted ballot boxes having the second lowest turnout, no law being in the middle, and broadly available having the second highest turnout. 







