---
title: "PS270 Final Project"
author: "Jared Haberman"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

  This project explores how voter accessibility can impact voter turnout, focusing specifically on ballot drop boxes. The research question is as follows, **"Does increasing the availability of ballot drop boxes increase voter participation in elections?"** My hypothesis is that **if ballot drop boxes are more available for voters, then there will by higher voter turnout.** The null hypothesis is that the availability of ballot drop boxes does not affect voter turnout. 
  This research is important in determining how to increase voter turnout. Less than two-thirds of eligible Americans voted in 2024, a number that can be greatly increased. Increasing voter turnout allows for American government to become more democratic, as more people are using their voice. If ballot drop boxes cause more people to vote, then they should be used across the country. 
  
## Data
  
  The predictive variable is state laws regarding ballot drop boxes. States can either require ballot drop boxes, allow them but not require them, have no law on drop boxes, restrict the use of ballot drop boxes, or ban their use altogether. These will be the categories in which the variable is organized. The outcome variable is in voter turnout. Voter turnout is measured by state. Voter turnout is the proportion of eligible voter who actually voted in the election. 
  If states that require ballot drop boxes have higher voter turnout, then my hypothesis will be supported. However, if states with more restrictive use on ballot drop boxes have higher voter turnout, or there is no correlation between ballot drop boxes and voter turnout, then my hypothesis will be disproved. Mathematically, the hypothesis will be tested at a 95% confidence interval. Therefore, I will calculate the p-value, and if it is less than 0.05, then the null hypothesis will be rejected. 


```{r}
library(tidyverse)
```

```{r}
ballot <- read.table(text = "
State Ballot_Box_Law Voter_Participation_2024
AL No_Law 58.7
AK No_Law 62.1
AZ Broadly_Available 61.7
AR Banned 52.8
CA Required 64.7
CO Required 70.1
CT Broadly_Available 66.3
DE Broadly_Available 69.8
FL Restricted 60.3
GA Restricted 64.4
HI Broadly_Available 58.5
ID Broadly_Available 65.2
IL Broadly_Available 65.9
IN Restricted 60.7
IA Restricted 71.8
KS No_Law 70.7
KY Required 67.7
LA Restricted 58.0
ME Broadly_Available 67.1
MD Required 71.1
MA Broadly_Available 68.9
MI Required 72.3
MN Broadly_Available 75.9
MS Banned 68.9
MO Banned 69.5
MT No_Law 69.2
NE Broadly_Available 67.4
NV Required 66.9
NH No_Law 71.9
NJ Required 72.5
NM Required 66.0
NY Broadly_Available 59.2
NC Restricted 64.6
ND Broadly_Available 71.2
OH Restricted 67.4
OK Banned 59.7
OR Required 75.3
PA Broadly_Available 69.7
RI Required 70.0
SC Restricted 62.7
SD Banned 58.9
TN Restricted 64.8
TX Banned 57.9
UT Required 66.9
VT Broadly_Available 71.7
VA Required 72.9
WA Required 70.0
WV Restricted 59.5
WI Broadly_Available 72.2
WY Broadly_Available 63.8
", header = TRUE) 
glimpse(ballot)
```

```{r}
ggplot(ballot, aes(x = State,
                   y = Voter_Participation_2024)) +
  geom_point() +
  labs(title = "Voter Participation by State",
       x = "State",
       y = "Voter Participation in 2024 (%)"
  )
```

This graph displays the different rates of voter participation per state. It displays how much voter participation can vary, with some states being close to 80% participation and some being close to 50% participation. We are looking to organize this according to ballot drop box availability, to see the impact it may have on voter participation. 

## Results

```{r}
ballot_summary <- ballot |>
  group_by(Ballot_Box_Law) |>
  summarize(
    avg_turnout = mean(Voter_Participation_2024, na.rm = TRUE),
    .groups = "drop"
  )

ballot_summary

```

```{r}
Ballot_Plot <- ballot_summary |>
  ggplot(mapping = aes(x = fct_reorder(Ballot_Box_Law, avg_turnout),
                       y = avg_turnout)) +
  geom_bar(stat = "identity") +
  labs(x = "Ballot Box Law",
       y = "Avg Voter Turnout",
       title = "How Ballot Box Laws Effect Voter Turnout")
Ballot_Plot
```

```{r}
ballot_summary_table <- ballot_summary |>
  rename("Ballot Box Availability" = Ballot_Box_Law,
         "Average Voter Turnout" = avg_turnout)

knitr::kable(ballot_summary_table, digits = 2)
```

This data shows a somewhat small, but noticeable effect of ballot box availability on voter turnout. The more supportive a law is regarding ballot boxes, the higher voter turnout is. In states where ballot boxes are banned, voter turnout is about 61.3%, while in states where ballot boxes are required, voter turnout is at 69.7%. The rest of the availability options follow this trend as well, with restricted ballot boxes having the second lowest turnout, no law being in the middle, and broadly available having the second highest turnout. 

```{r}
fit <- lm(Voter_Participation_2024 ~ Ballot_Box_Law, data = ballot)
fit

summary(fit)

library(broom)
tidy(fit)

```




The regression analysis works with ballot boxes being banned as the intercept. When ballot boxes are banned, the average voter turnout per state is 61.28%. This result is statistically significant at 95% confidence as the p-value is between 0 and 0.001. The coefficient for ballot boxes being broadly available is 5.87. This means that on average, in states where ballot drop boxes are available, voter turnout is about 5.87 percentage points higher than in states where ballot drop boxes are banned. This result is statistically significant as the p-value is less than 0.05. The coefficient for states with no ballot box law is 5.24. This means on average their voter turnout is 5.23 percentage points higher than states where ballot drop boxes are banned. This is not statistically significant however as the p-value is greater than 0.05. The coefficient for states where ballot drop boxes are required is 8.44. This means that states where ballot drop boxes are required have on average 8.44 percentage point higher turnout than states where ballot drop boxes are banned. This result is statistically significant as the p-value is less than 0.05. Finally, the coefficient for ballot drop boxes being restricted is 2.14. This means that on average, states that restrict ballot drop box use have on average 2.14 percentage point higher voter turnout than states where ballot drop boxes are banned. 

It makes sense that the regression found no statistical significance for no ballot box law and ballot boxes being restricted. When relative to ballot boxes being banned, ballot boxes being restricted is not much different. 




